---
title: "BSIT 375: Multivariate Statistical Analysis - Lab Components"
author: "Instructor: Tek Raj Pant"
date: "March 2026"
output: html_document
---

```{r setup, include=FALSE}
required_packages <- c( "caret", "infotheo", "knitr", "kableExtra")

# Check which packages are not installed and install them
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load all packages
lapply(required_packages, library, character.only = TRUE)

```

## Two-Sample t-Test: App Redesign A/B Test

**Scenario:** Comparing daily active usage (minutes) between Group A (Old) and Group B (New).

```{r}
group_a <- rnorm(40, mean = 25.3, sd = 5.1)
group_b <- rnorm(35, mean = 28.7, sd = 6.3)

# Conduct the test
t_result <- t.test(group_b, group_a, alternative = "two.sided", var.equal = FALSE)

(t_result)

```

## Two-Sample Z-Test: Email Marketing Proportions

**Scenario:** Comparing open rates for Subject Line A vs. Subject Line B.

```{r}
# Data from lecture notes
successes <- c(325, 412)
totals <- c(5000, 4800)

# prop.test performs the chi-square test for proportions 
# (which is equivalent to the z-test squared)
prop_result <- prop.test(successes, totals, correct = FALSE)

prop_result
```

## Chi-Square Test: Customer Satisfaction Homogeneity

**Scenario:** Testing if satisfaction levels are consistent across North, South, and West regions.

```{r}
# Constructing the contingency table
satisfaction_data <- matrix(c(120, 60, 20, 
                              90, 50, 30, 
                              110, 40, 25), 
                            nrow = 3, byrow = TRUE)

rownames(satisfaction_data) <- c("North", "South", "West")
colnames(satisfaction_data) <- c("Satisfied", "Neutral", "Dissatisfied")

# Perform Chi-Square Test
chi_result <- chisq.test(satisfaction_data)

print(chi_result)
# Access expected frequencies calculated in the lecture notes
chi_result$expected
```

## Chi-Square Goodness of Fit: Website Traffic

**Scenario:** Comparing observed traffic sources against a predicted 50/30/15/5 distribution.

```{r}
observed <- c(Direct = 480, Search = 320, Social = 140, Referral = 60)
expected_probs <- c(0.50, 0.30, 0.15, 0.05)

gof_result <- chisq.test(observed, p = expected_probs)

print(gof_result)
```

## ANOVA: Marketing Campaign Comparison

**Scenario:** Weekly sales across three different campaigns.\

```{r}
# Create tidy data frame
campaign_data <- data.frame(
  Campaign = rep(c("A", "B", "C"), each = 5),
  Sales = c(12, 14, 11, 13, 10,  # A
            15, 16, 14, 18, 17,  # B
            10, 12, 11, 9, 13)   # C
)

# Fit ANOVA model
anova_model <- aov(Sales ~ Campaign, data = campaign_data)
summary(anova_model)

# Post-hoc analysis (Tukey HSD)
TukeyHSD(anova_model)
```

## Simple Linear Regression: Study Hours

**Scenario:** Predicting exam scores based on hours studied.

```{r}
# 1. Define the data
study_data <- data.frame(
  Hours = c(2, 3, 4, 5, 6),
  Score = c(65, 70, 80, 85, 90)
)

# 2. Fit the OLS model
reg_model <- lm(Score ~ Hours, data = study_data)

# 3. Create the scatter plot
plot(study_data$Hours, study_data$Score, 
     pch = 16,            # Solid circle point style
     col = "blue",        # Point color
     main = "Regression: Study Hours vs. Exam Score",
     xlab = "Hours", 
     ylab = "Score")

# 4. Add the regression line
# abline() can take the model object directly to draw the best-fit line
abline(reg_model, col = "red", lwd = 2)
```

## Multiple Regression: House Prices

**Scenario:** Predicting price based on size and number of bedrooms.

```{r}
house_data <- data.frame(
  Price = c(2.5, 3.0, 2.8, 3.5, 3.2),
  Size = c(1.5, 2.0, 1.8, 2.2, 2.0),
  Bedrooms = c(3, 4, 3, 4, 3)
)

# Fit multiple regression
multi_model <- lm(Price ~ Size + Bedrooms, data = house_data)

summary(multi_model)

# Check for Multicollinearity (VIF)
# requires 'car' package: install.packages("car")
# car::vif(multi_model)
```
