---
title: "R Notebook"
output: html_notebook
---

## Question 1: Estimating Cloud Storage Latency (Z-Interval)

**Problem:** A data engineer wants to estimate the average time it takes to retrieve a file from a specific AWS S3 bucket. Based on previous logs, the population standard deviation is known to be $\sigma = 12$ ms. A sample of $n = 100$ retrievals shows a mean time of $\bar{x} = 45$ ms.

-   Calculate a **95% Confidence Interval** for the true mean retrieval time.

    ```{r}
    # Parameters
    x_bar <- 45
    sigma <- 12
    n <- 100
    conf_level <- 0.95

    # Calculations
    z_crit <- qnorm(1 - (1 - conf_level)/2)
    se <- sigma / sqrt(n)
    me <- z_crit * se

    # Result
    ci <- c(x_bar - me, x_bar + me)
    cat("95% CI for Mean Latency:", round(ci[1], 2), "ms to", round(ci[2], 2), "ms")
    ```

    ## Question 2: Testing API Performance (t-test)

    **Problem:** A developer claims that a new microservice averages less than **200 ms** per request. You run a small pilot test of $n = 15$ requests and find a sample mean of $\bar{x} = 188$ ms and a sample standard deviation of $s = 24$ ms.

    -   At a **5% significance level (**$\alpha = 0.05$), is there enough evidence to support the developer's claim? ($H_a: \mu < 200$)

    ```{r}
        # Summary Statistics
        mu_0 <- 200
        x_bar <- 188
        s <- 24
        n <- 15
        alpha <- 0.05

        # Manual t-test
        t_stat <- (x_bar - mu_0) / (s / sqrt(n))
        p_val <- pt(t_stat, df = n - 1) # Lower tail because claim is "less than"

        # Result
        cat("t-statistic:", round(t_stat, 3), "\np-value:", round(p_val, 4))
        if(p_val < alpha) {
          print("Conclusion: Reject H0. Evidence supports the developer's claim.")
        } else {
          print("Conclusion: Fail to Reject H0. Evidence does not support the claim.")
        }
    ```

## Question 3: User Engagement Rates (Proportion CI)

**Problem:** A social media app wants to estimate the proportion of users who click on a new "reels" feature. In a random sample of $n = 1000$ active users, $220$ clicked the feature.

-   Calculate a **99% Confidence Interval** for the true population proportion of users who click the feature.

    ```{r}
    # Parameters
    n <- 1000
    x <- 220
    p_hat <- x / n
    conf_level <- 0.99

    # Built-in R function (correct=FALSE for standard normal approximation)
    prop_result <- prop.test(x, n, conf.level = 0.99, correct = FALSE)

    # Result
    cat("Sample Proportion:", p_hat, "\n99% CI for Proportion:", 
        round(prop_result$conf.int[1], 4), "to", round(prop_result$conf.int[2], 4))
    ```
