---
title: "BSIT 375: Knowledge Discovery and Data Science"
subtitle: "Lecture Notes & Lab Supplement: The Data Preprocessing Pipeline"
author: "Prepared by: Tek Raj Pant"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
---

```{r setup, include=TRUE}
required_packages <- c("tidyverse", "caret", "infotheo", "knitr", "kableExtra")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

# Set seed for reproducibility
set.seed(123)


```

# Pillar 1: Data Cleaning

## Handling Missing Data

1.  **MCAR (Missing Completely at Random)**: $P(M | Y_{obs}, Y_{mis}) = P(M)$. Purely accidental; no relationship with any value.

```{r}
# Initializing sample data
df_raw <- data.frame(
  ID = 1:5,
  Score = c(85, 90, NA, 88, NA)
)
kable(df_raw, caption = "Initial data")
# Handling MCAR: Listwise Deletion (Removing any row with NA)
df_mcar_fixed <- df_raw %>% 
  drop_na(Score)

kable(df_mcar_fixed, caption = "MCAR Treatment: Listwise Deletion")
```

2.  **MAR (Missing at Random)**: $P(M | Y_{obs}, Y_{mis}) = P(M | Y_{obs})$. Systematic but observable through other variables.

```{r}
df_mar <- data.frame(
  Age = c(25, 30, NA, 22, 28, NA, 35),
  Salary = c(50000, 60000, 55000, 48000, 58000, 62000, 70000)
)

# Handling MAR: Simple Imputation (Replacing NA with the Mean)
df_mar_fixed <- df_mar %>%
  mutate(Age = replace_na(Age, mean(Age, na.rm = TRUE)))

kable(df_mar_fixed, caption = "MAR Treatment: Mean Imputation")
```

3.  **MNAR (Missing Not at Random)**: Depends on $Y_{mis}$. The data is "hiding" (e.g., high-income earners not reporting salary).

```{r}
df_mnar <- data.frame(
  User = c("A", "B", "C", "D"),
  Gender = c("M", "F", NA, "M") 
)

# Handling MNAR: Flagging as a new category "Unknown"
df_mnar_fixed <- df_mnar %>%
  mutate(Gender = as.character(Gender)) %>%
  mutate(Gender = replace_na(Gender, "Not Disclosed"))


kable(df_mnar_fixed, caption = "MNAR Treatment: Category Flagging")
```

## Noise Smoothing and Outlier Detection

Outlier detection identifies data points that deviate significantly from the rest. Two common statistical heuristics are the Z-Score method (parametric) and the IQR method (non-parametric).

### Z-Score Method (Parametric)

Assumes a Normal Distribution. Flag $x_i$ if $|z_i| > 3$.

$$ z_i = \frac{x_i - \mu}{\sigma} $$

```{r zscore}
# Generate data with outliers
set.seed(123)
data_z <- c(rnorm(50, mean = 50, sd = 5), 95, 10)  # two outliers added

# Calculate Z-scores
z_scores <- (data_z - mean(data_z)) / sd(data_z)
outliers_z <- data_z[abs(z_scores) > 3]

# Create a data frame for display
z_df <- data.frame(
  Value = data_z,
  Z_Score = round(z_scores, 2),
  Outlier = ifelse(abs(z_scores) > 3, "Yes", "No")
)
kable(z_df[abs(z_scores) > 3 | seq_along(data_z) %in% 51:52, ], 
      caption = "Z-Score Outlier Detection (showing outliers and nearby values)")

cat("Z-Score Outliers detected:", outliers_z)
```

### IQR Method (Non-Parametric)

Uses the interquartile range and is robust to non-normal distributions.

$$ IQR = Q_3 - Q_1 $$ Inner fences: $[Q_1 - 1.5 \times IQR, Q_3 + 1.5 \times IQR]$

```{r iqr}
# Using the same data
Q1 <- quantile(data_z, 0.25)
Q3 <- quantile(data_z, 0.75)
IQR_val <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR_val
upper_bound <- Q3 + 1.5 * IQR_val

outliers_iqr <- data_z[data_z < lower_bound | data_z > upper_bound]

# Create a summary
iqr_df <- data.frame(
  Value = data_z,
  Outlier_IQR = ifelse(data_z < lower_bound | data_z > upper_bound, "Yes", "No")
)
kable(iqr_df[data_z < lower_bound | data_z > upper_bound | seq_along(data_z) %in% 51:52, ],
      caption = "IQR Outlier Detection")

cat("IQR Outliers detected:", outliers_iqr)

# Visual comparison
par(mfrow = c(1, 2))
boxplot(data_z, main = "Boxplot (IQR Outliers)", col = "lightblue")
hist(data_z, main = "Histogram with Z-score outliers", col = "lightgreen")
abline(v = c(mean(data_z) - 3*sd(data_z), mean(data_z) + 3*sd(data_z)), col = "red", lwd = 2)
```

## Redundancy and Correlation Analysis

Redundancy occurs when an attribute can be derived from another, leading to multi-collinearity. We use different measures for numerical and categorical data.

### Pearson Correlation Coefficient (Numerical Data)

Measures linear relationship between two continuous variables.

$$ r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{(n-1)s_x s_y} $$

```{r pearson}
# 1. Setup Data
N <- 100
set.seed(123)

# Independent variable
x_val <- seq(1, 10, length.out = N)

# Dependent variables for different scenarios
y_positive <- x_val                         # Perfect Positive
y_negative <- 10 - x_val + rnorm(N, sd=0.5) # Strong Negative
y_none     <- rnorm(N)                     # No Correlation

# 2. Create the Layout (1 row, 3 columns)
par(mfrow = c(1, 3))

# Plot 1: Positive
plot(x_val, y_positive, main="Positive (r=1.0)", 
     xlab="Variable A", ylab="Variable B", col="blue", pch=19)
abline(lm(y_positive ~ x_val), col="red") # Trendline

# Plot 2: Negative
plot(x_val, y_negative, main="Negative (r ≈ -0.9)", 
     xlab="Variable A", ylab="Variable B", col="darkgreen", pch=19)
abline(lm(y_negative ~ x_val), col="red") # Trendline

# Plot 3: None
plot(x_val, y_none, main="None (r ≈ 0)", 
     xlab="Variable A", ylab="Variable B", col="gray", pch=19)
abline(lm(y_none ~ x_val), col="red") # Trendline

# Reset layout to default
par(mfrow = c(1, 1))
```

### Chi-Square ($\chi^2$) Test (Categorical Data)

Tests independence between two categorical variables. High dependence implies redundancy.

$$ \chi^2 = \sum \frac{(O_{ij} - E_{ij})^2}{E_{ij}} $$ where $E_{ij} = \frac{(\text{row total}_i \times \text{column total}_j)}{\text{grand total}}$

```{r chisq}
# Contingency table: Gender vs. Shopping Preference
observed <- matrix(c(20, 30, 40, 10), nrow = 2, byrow = TRUE,
                   dimnames = list(Gender = c("Male", "Female"),
                                   Preference = c("Online", "Offline")))
observed

# Chi-square test
chisq_test <- chisq.test(observed)

# Display results
kable(observed, caption = "Observed Frequencies")
kable(chisq_test$expected, caption = "Expected Frequencies under Independence")

cat("Chi-square statistic:", chisq_test$statistic, "\n")
cat("p-value:", chisq_test$p.value, "\n")
cat("Degrees of freedom:", chisq_test$parameter, "\n")

# Manual calculation to verify
expected <- chisq_test$expected
chi_manual <- sum((observed - expected)^2 / expected)
cat("Manual chi-square:", chi_manual, "\n")
```

# Pillar 2: Data Transformation

Data transformation converts data into a state that enhances model performance. This includes scaling and encoding.

## Normalization Techniques

### Min-Max Scaling (Rescaling)

Scales data to a fixed range, typically $[0,1]$.

$$ x' = \frac{x - \min(x)}{\max(x) - \min(x)} $$

```{r minmax}
# Sample data with an outlier
data_scale <- c(rnorm(10, mean = 100, sd = 20), 500)

# Min-Max scaling function
minmax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

scaled_minmax <- minmax(data_scale)

# Create comparison table
df_scale <- data.frame(
  Original = data_scale,
  MinMax = round(scaled_minmax, 4)
)
kable(df_scale, caption = "Min-Max Scaling (with outlier)")
```

### Standardization (Z-score Normalization)

Centers data to mean 0 and standard deviation 1.

$$ z = \frac{x - \mu}{\sigma} $$

```{r standardization}
scaled_z <- as.numeric(scale(data_scale))

df_scale$Z_score <- round(scaled_z, 4)
kable(df_scale, caption = "Standardization (Z-score) with outlier")

# Compare robustness
cat("Min-Max scaled outlier:", scaled_minmax[which.max(data_scale)], "\n")
cat("Z-score of outlier:", scaled_z[which.max(data_scale)], "\n")
```

## Categorical Encoding

### One-Hot Encoding (Nominal Data)

Creates binary columns for each category.

```{r onehot}
# Sample nominal data
colors <- c("Red", "Blue", "Green", "Blue", "Red", "Green")

# Using model.matrix for one-hot encoding
onehot_matrix <- model.matrix(~ colors - 1)  # -1 removes intercept
colnames(onehot_matrix) <- gsub("colors", "", colnames(onehot_matrix))

kable(onehot_matrix, caption = "One-Hot Encoded Colors")

# Avoiding dummy variable trap: drop one column
onehot_reduced <- onehot_matrix[, -1, drop = FALSE]
kable(onehot_reduced, caption = "One-Hot Encoding (k-1 columns)")
```

### Label Encoding (Ordinal Data)

Assigns integers to ordered categories.

```{r label-encoding}
# Ordinal data: Education level
education <- c("High School", "Bachelor", "Master", "PhD", "Bachelor", "High School")
# Define order
edu_order <- c("High School", "Bachelor", "Master", "PhD")
edu_encoded <- as.numeric(factor(education, levels = edu_order, ordered = TRUE))

df_edu <- data.frame(
  Education = education,
  Encoded = edu_encoded
)
kable(df_edu, caption = "Label Encoding for Ordinal Data")
```

| Feature        | One-Hot Encoding        | Label Encoding        |
|----------------|-------------------------|-----------------------|
| Data Type      | Nominal (No Order)      | Ordinal (Has Order)   |
| Dimensionality | Increases ($k$ columns) | Stays same (1 column) |
| Algorithm Risk | Multi-collinearity      | Improper Ordinal Bias |

# Pillar 3: Data Reduction

Data reduction techniques reduce data volume while preserving integrity. This improves efficiency and mitigates the curse of dimensionality.

## Dimensionality Reduction: Principal Component Analysis (PCA)

PCA transforms correlated variables into uncorrelated principal components that capture maximum variance.

```{r pca}
# Create correlated dataset
set.seed(123)
n <- 50
x1 <- rnorm(n, mean = 10, sd = 3)
x2 <- 0.7 * x1 + rnorm(n, sd = 2)   # correlated with x1
x3 <- rnorm(n, mean = 5, sd = 1)     # independent
x4 <- -0.5 * x1 + 0.3 * x2 + rnorm(n, sd = 1)

df_pca <- data.frame(x1, x2, x3, x4)

# Perform PCA (scale = TRUE to standardize)
pca_result <- prcomp(df_pca, center = TRUE, scale. = TRUE)

# Summary of variance explained
summary(pca_result)

# Variance table
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
var_table <- data.frame(
  PC = paste0("PC", 1:4),
  `Variance` = round(var_explained, 3),
  `Cumulative` = round(cumsum(var_explained), 3)
)
kable(var_table, caption = "Proportion of Variance Explained by Principal Components")

# Scree plot
plot(pca_result, type = "l", main = "Scree Plot")

# Biplot
biplot(pca_result, main = "PCA Biplot")
```

## Numerosity Reduction

### Parametric Reduction: Linear Regression

Instead of storing all data, store only model parameters.

```{r regression}
x <- 1:10
y <- 3 * x + rnorm(10, sd = 2)
model <- lm(y ~ x)

# Model parameters (intercept and slope)
coef(model)

# Plot
plot(x, y, main = "Linear Regression Model", pch = 16)
abline(model, col = "red", lwd = 2)
```

### Non-Parametric Reduction: Histograms, Clustering, Sampling

```{r reduction-examples}
# Histogram binning (reduce continuous data to bin counts)
hist_data <- rnorm(200, mean = 50, sd = 10)
hist(hist_data, breaks = 10, col = "lightblue", main = "Histogram Binning")

# Clustering: represent data by cluster centroids
kmeans_result <- kmeans(df_pca, centers = 3, nstart = 10)
centroids <- kmeans_result$centers
kable(centroids, caption = "Cluster Centroids (Reduced Representation)")

# Sampling
sample_rows <- sample(1:nrow(df_pca), size = 10)
sample_data <- df_pca[sample_rows, ]
kable(sample_data, caption = "Random Sample (10 rows)")
```

| Strategy       | Approach                 | Key Benefit                     |
|------------------|------------------------|-------------------------------|
| Dimensionality | Feature Extraction (PCA) | Eliminates redundant attributes |
| Numerosity     | Data Approximation       | Reduces storage and CPU load    |
| Sampling       | Representative Subset    | Faster prototyping and EDA      |

# Pillar 4: Data Discretization

Discretization converts continuous variables into interval labels, simplifying data for rule-based algorithms.

## Binning Techniques

### Equal-Width (Distance) Binning

Divides the range into intervals of equal width.

```{r equalwidth}
# Example prices
prices <- c(5, 10, 11, 13, 15, 35, 50, 55, 72, 92, 204, 215)

# Equal-width binning with 3 bins
breaks_width <- seq(min(prices), max(prices), length.out = 4)  # 3 bins need 4 breaks
bins_width <- cut(prices, breaks = breaks_width, include.lowest = TRUE,
                  labels = c("Bin1 [5,75]", "Bin2 (75,145]", "Bin3 (145,215]"))

df_width <- data.frame(Price = prices, Bin = bins_width)
kable(df_width, caption = "Equal-Width Binning (3 bins)")
```

### Equal-Frequency (Depth) Binning

Ensures each bin contains approximately the same number of observations.

```{r equalfreq}
# Equal-frequency binning using quantiles
bins_freq <- cut(prices, 
                 breaks = quantile(prices, probs = seq(0, 1, length.out = 4)),
                 include.lowest = TRUE,
                 labels = c("Bin1 [5,13]", "Bin2 (13,55]", "Bin3 (55,215]"))

df_freq <- data.frame(Price = prices, Bin = bins_freq)
kable(df_freq, caption = "Equal-Frequency Binning (3 bins)")
```

## Concept Hierarchies

Group low-level data into higher-level concepts using domain knowledge.

```{r concept-hierarchy}
# Age groups
ages <- c(12, 25, 70, 8, 30, 65, 18, 45, 80, 5)

# Define breaks and labels
age_groups <- cut(ages, 
                  breaks = c(0, 18, 60, 120),
                  labels = c("Child", "Adult", "Senior"),
                  right = FALSE)

df_age <- data.frame(Age = ages, Group = age_groups)
kable(df_age, caption = "Age Group Hierarchy")
```

| Method | Optimization Goal | Risk |
|------------------------|------------------------|------------------------|
| Equal-Width | Maintains value distribution | Sensitive to outliers |
| Equal-Frequency | Maintains sample distribution | May group very different values |
| Entropy-Based | Maximizes Class Purity | Requires target labels (Supervised) |

# 
