---
title: "BSIT 375: Knowledge Discovery and Data Science - Lecture Notes: Exploratory Data Analysis"
author: "Prepared by: Tek Raj Pant"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
```

```{r getting-to-know-data}
# Use built-in dataset 'mtcars' for illustration
data(mtcars)
str(mtcars)           # structure and data types
dim(mtcars)           # dimensions
summary(mtcars)       # basic summaries for all variables

# Check for missing values
anyNA(mtcars)
```

------------------------------------------------------------------------

## Exploring Categorical Variables

Categorical variables take values from a set of categories. EDA for categorical variables focuses on frequencies and proportions.

### Frequency Tables

A frequency table lists each category along with its count and percentage. For example, a survey might have gender categories: Male (45%), Female (53%), Other (2%).

```{r freq-tables}
# Create a small categorical dataset
survey <- data.frame(
  gender = c("Male", "Female", "Female", "Male", "Other", "Female", "Male", "Female")
)

# Frequency table
freq <- table(survey$gender)
freq

# Proportions
prop.table(freq)
```

### Bar Charts and Pie Charts

Bar charts display frequencies as bar heights; pie charts show proportions as slices. These visualizations quickly reveal dominant categories and rare ones.

```{r bar-pie}
# Bar chart using base R
barplot(freq, main = "Gender Distribution", xlab = "Gender", ylab = "Count", col = "steelblue")

# Pie chart
pie(freq, main = "Gender Distribution", col = c("lightblue", "pink", "lightgreen"))

# Using ggplot2 for a more polished bar chart
ggplot(survey, aes(x = gender)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count")
```

### Contingency Tables (Cross-Tabulation)

When two categorical variables are of interest, a contingency table shows the joint distribution. For example, a table of gender versus purchase (yes/no) can suggest associations.

```{r contingency}
# Add a purchase column
survey$purchase <- c("Yes", "No", "Yes", "No", "Yes", "No", "No", "Yes")

# Contingency table
cont_table <- table(survey$gender, survey$purchase)
cont_table

# Mosaic plot for visual association
mosaicplot(cont_table, main = "Gender vs Purchase", color = TRUE)
```

### Chi-Square Test for Independence

Although more formal, the chi-square test can be used during EDA to assess whether two categorical variables are related. The test statistic is: $$
\chi^2 = \sum \frac{(O_{ij} - E_{ij})^2}{E_{ij}}
$$ where $O_{ij}$ is the observed count and $E_{ij}$ is the expected count under independence.

```{r chisq}
chisq.test(cont_table)
```

------------------------------------------------------------------------

## Exploring Numeric Variables

Numeric variables require different tools. The goal is to understand their distribution, central tendency, spread, and shape.

### Summary Statistics

-   **Center**: mean, median, mode.
-   **Spread**: range, variance, standard deviation, interquartile range (IQR).
-   **Shape**: skewness (symmetric, left‑skewed, right‑skewed) and kurtosis (peakedness).

```{r numeric-summary}
# Use mtcars$mpg
mpg <- mtcars$mpg

mean(mpg)
median(mpg)
# Mode function (R doesn't have a built-in mode for numeric)
mode_mpg <- as.numeric(names(sort(table(mpg), decreasing = TRUE)[1]))
mode_mpg

range(mpg)
var(mpg)
sd(mpg)
IQR(mpg)

# Skewness and kurtosis (using moments package)
# install.packages("moments") if needed
library(moments)
skewness(mpg)
kurtosis(mpg)
```

### Histograms

A histogram partitions the data into bins and shows the frequency of observations in each bin. It reveals modality (unimodal, bimodal), gaps, and outliers.

```{r hist}
hist(mpg, breaks = 10, col = "lightblue", main = "Histogram of MPG", xlab = "Miles per Gallon")

# ggplot2 version
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "black") +
  labs(title = "Histogram of MPG", x = "Miles per Gallon")
```

### Boxplots

A boxplot displays the five‑number summary: minimum, first quartile (Q1), median, third quartile (Q3), and maximum. Points beyond $Q1 - 1.5 \times IQR$ or $Q3 + 1.5 \times IQR$ are plotted individually as potential outliers.

```{r boxplot}
boxplot(mpg, main = "Boxplot of MPG", ylab = "Miles per Gallon", col = "lightgreen")

# Horizontal boxplot
boxplot(mpg, horizontal = TRUE, main = "Boxplot of MPG", xlab = "Miles per Gallon", col = "lightgreen")
```

### Quantile-Quantile (Q-Q) Plots

Q-Q plots compare the quantiles of the data against the quantiles of a theoretical distribution (e.g., normal). If the points fall roughly on a straight line, the data follow that distribution.

```{r qqplot}
qqnorm(mpg, main = "Normal Q-Q Plot of MPG")
qqline(mpg, col = "red")
```

------------------------------------------------------------------------

## Exploring Multivariate Relationships

Understanding relationships between variables is central to EDA.

### Scatter Plots

For two numeric variables, a scatter plot shows each observation as a point. Patterns (linear, curved, clustered) can be detected. Color or size can encode a third variable.

```{r scatter}
plot(mtcars$wt, mtcars$mpg, 
     xlab = "Weight (1000 lbs)", ylab = "Miles per Gallon", 
     main = "Scatter Plot: MPG vs Weight", pch = 19, col = "blue")

# ggplot2 version with a third variable (cyl) as color
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "MPG vs Weight", x = "Weight", y = "MPG", color = "Cylinders")
```

### Correlation

The Pearson correlation coefficient $r$ measures the strength and direction of a linear relationship: $$
r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}
$$ Values near $\pm 1$ indicate strong linear association; 0 indicates no linear relationship. However, scatter plots are essential because Anscombe's quartet shows that different patterns can yield the same correlation.

```{r cor}
cor(mtcars$wt, mtcars$mpg)

# Correlation matrix for several variables
cor_matrix <- cor(mtcars[, c("mpg", "disp", "hp", "wt")])
round(cor_matrix, 2)
```

### Side-by-Side Boxplots

To compare a numeric variable across categories, side‑by‑side boxplots are invaluable. For example, income distribution by education level.

```{r side-by-side-box}
# Using mtcars: compare mpg across number of cylinders
boxplot(mpg ~ cyl, data = mtcars, 
        xlab = "Number of Cylinders", ylab = "Miles per Gallon",
        main = "MPG by Cylinder Count", col = c("lightblue", "lightgreen", "lightpink"))

# ggplot2 version
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot() +
  labs(title = "MPG by Cylinder Count", x = "Cylinders", y = "MPG") +
  theme(legend.position = "none")
```

### Heatmaps

A heatmap of correlations (or other measures) provides a quick overview of many variables. Color intensity represents the strength of association.

```{r heatmap}
# Simple heatmap using base R's image()
image(cor_matrix, main = "Correlation Heatmap", axes = FALSE)
axis(1, at = seq(0,1,length.out = ncol(cor_matrix)), labels = colnames(cor_matrix), las = 2)
axis(2, at = seq(0,1,length.out = ncol(cor_matrix)), labels = rownames(cor_matrix), las = 2)

# Using corrplot package for better visualization
# install.packages("corrplot") if needed
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black")
```

------------------------------------------------------------------------

## Selecting Interesting Subsets of the Data for Further Investigation

EDA often reveals subgroups that behave differently. For instance, a scatter plot might show two distinct clusters. These subsets can be isolated for separate analysis, leading to more targeted models. Techniques include:

-   **Brushing** in interactive graphics.
-   **Conditioning** (trellis plots) where plots are drawn for each level of a conditioning variable.
-   **Automatic interaction detection** (e.g., regression trees) that partitions the data.

```{r subsets}
# Example: subset of cars with high mpg (> 25)
high_mpg <- subset(mtcars, mpg > 25)
high_mpg

# Conditioning plot: scatter plot by cylinder
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl) +
  labs(title = "MPG vs Weight by Number of Cylinders")
```

------------------------------------------------------------------------

## Using EDA to Uncover Anomalous Fields

Anomalous fields are variables that contain errors or unexpected values. EDA can uncover:

-   **Implausible values**: Age = 200, negative income.
-   **Inconsistent coding**: "M" and "Male" in the same variable.
-   **Unexpected missing patterns**: Missing not at random.

Graphical tools like histograms and frequency tables help spot these issues. For example, a histogram of age with a spike at 999 likely indicates a placeholder for missing.

```{r anomalies}
# Create a small dataset with an anomaly
data_with_error <- mtcars
data_with_error$mpg[5] <- 999   # introduce an extreme value

# Histogram shows spike at 999
hist(data_with_error$mpg, breaks = 20, main = "MPG with Anomaly", xlab = "MPG")

# Boxplot highlights outlier
boxplot(data_with_error$mpg, main = "Boxplot with Anomaly")

# Frequency table of categorical variable with inconsistent coding
survey_bad <- data.frame(
  gender = c("Male", "Female", "M", "F", "male", "female")
)
table(survey_bad$gender)   # shows multiple categories for same gender
```

------------------------------------------------------------------------

## Binning Based on Predictive Value

Binning (discretizing) a numeric variable can improve interpretability and handle non‑linear relationships. However, arbitrary binning may lose information. An alternative is to bin based on the variable's relationship with a target (predictive value). For instance, using a decision tree to find optimal cut points that maximize separation of classes. This yields bins that are meaningful for prediction.

Common approaches:

-   **Equal width bins**: Simple but may hide patterns.
-   **Equal frequency bins**: Each bin has roughly the same number of observations.
-   **Supervised binning**: Use the target variable to define bins (e.g., chi‑merge, entropy‑based).

```{r binning}
# Equal width bins for mpg
mtcars$mpg_eqwidth <- cut(mtcars$mpg, breaks = 3, labels = c("Low", "Medium", "High"))
table(mtcars$mpg_eqwidth)

# Equal frequency bins
mtcars$mpg_eqfreq <- cut(mtcars$mpg, 
                         breaks = quantile(mtcars$mpg, probs = seq(0,1,length.out=4)), 
                         include.lowest = TRUE, 
                         labels = c("Low", "Medium", "High"))
table(mtcars$mpg_eqfreq)

# Supervised binning using a decision tree to predict 'am' (transmission) from mpg
library(rpart)
tree <- rpart(am ~ mpg, data = mtcars, method = "class")
print(tree)   # shows splits at certain mpg values
```

------------------------------------------------------------------------

## Deriving New Variables: Flag Variables

Flag (indicator) variables are binary (0/1) variables that encode a condition. They can be derived from existing variables to capture important distinctions. Examples:

-   From a date variable, create flags for weekend/weekday, holiday, season.
-   From a numeric variable, create a flag for missing values.
-   From a categorical variable, create flags for each category (one‑hot encoding) or for rare categories combined as "other".

Flags can capture non‑linear effects and interactions without changing the scale.

```{r flag-vars}
# Create a flag for cars with high horsepower (> 200)
mtcars$high_hp_flag <- ifelse(mtcars$hp > 200, 1, 0)

# Flag for missing values (example: set some NAs)
mtcars_na <- mtcars
mtcars_na$mpg[1:3] <- NA
mtcars_na$missing_mpg_flag <- ifelse(is.na(mtcars_na$mpg), 1, 0)

# One-hot encoding for cylinder variable
mtcars$cyl_4 <- ifelse(mtcars$cyl == 4, 1, 0)
mtcars$cyl_6 <- ifelse(mtcars$cyl == 6, 1, 0)
mtcars$cyl_8 <- ifelse(mtcars$cyl == 8, 1, 0)
head(mtcars[, c("cyl", "cyl_4", "cyl_6", "cyl_8")])
```

------------------------------------------------------------------------

## Deriving New Variables: Numerical Variables

New numerical variables can be created by transforming or combining existing ones. Common derivations include:

-   **Ratios**: e.g., debt-to-income ratio.
-   **Differences**: e.g., year-over-year change.
-   **Aggregations**: e.g., average purchase amount per customer.
-   **Polynomial terms**: $x^2$, $x^3$ to model curvature.
-   **Interactions**: product of two variables to capture joint effects.

These derived variables can improve model performance by encoding domain knowledge.

```{r numeric-deriv}
# Ratio: weight per horsepower
mtcars$wt_per_hp <- mtcars$wt / mtcars$hp

# Polynomial term: squared weight
mtcars$wt_sq <- mtcars$wt^2

# Interaction: weight * horsepower
mtcars$wt_hp_interact <- mtcars$wt * mtcars$hp

head(mtcars[, c("wt", "hp", "wt_per_hp", "wt_sq", "wt_hp_interact")])
```
